---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "50%"
)
options(width=400)
```

# totalize

<!-- badges: start -->
<!-- badges: end -->

[Japanese README is here.](README-ja.md)

Aggregate values by groups and make their total then put into a matrix table

## Installation

You can install `totalize` from GitHub.

``` r
devtools::install_github("ichiromurata/totalize")
```

# How this works

Aggregating data can be done by ordinary R methods.

Take `datasets::CO2` for example. You can count the data by `table()`,

```{r}
table(interaction(datasets::CO2$Type, datasets::CO2$Treatment))
```

or aggregate values using`by()`,

```{r}
by(datasets::CO2$uptake, datasets::CO2[c("Type", "Treatment")], sum)
```

or `stats::aggregate()`.

```{r}
stats::aggregate(datasets::CO2["uptake"], datasets::CO2[c("Type", "Treatment")], sum)
```

But if you need subtotals and total (total "Type", total "Treatment" and both summed in the examples above), 
you have to calculate them through several steps since it involves multiple counting of the same data essentially 
and can't be done by one method execution.

`totalize` rolls up these operations.

```{r}
library(totalize)
totalize(datasets::CO2, c(Type, Treatment), val=uptake, asDF=TRUE)
```

Further, it provides matrix representation.

```{r}
totalize(datasets::CO2, row=Type, col=Treatment, val=uptake)
```

3 or more variables can be put into a (2d) matrix.

```{r}
totalize(datasets::CO2, row=Type, col=c(Treatment, conc), val=uptake)

totalize(datasets::CO2, row=c(Type, conc), col=Treatment, val=uptake)
```

Would you like to ignore NAs in the data when `FUN=sum`? Pass `na.rm=TRUE` to the arguments.

```{r}
# Missing values turn the totals into NA
transform(datasets::CO2, uptake_na=replace(uptake, c(1, 10), NA)) |>
    totalize(row=Type, col=c(Treatment, conc), val=uptake_na)

# Ignoring NAs
transform(datasets::CO2, uptake_na=replace(uptake, c(1, 10), NA)) |>
    totalize(row=Type, col=c(Treatment, conc), val=uptake_na, na.rm=TRUE)
```

## Single row or column matrix

`totailze` can make a 1-column matrix.

```{r}
totalize(datasets::CO2, row=c(Type, Treatment), val=uptake)
```

If you need a 1-row matrix, use transpose function `t()`.

```{r}
totalize(datasets::CO2, row=c(Type, Treatment), val=uptake) |> t()
```

# Rounding numbers

This package includes a rounding method which rounds half [away from zero](https://en.wikipedia.org/wiki/Rounding#Round_half_away_from_zero).

There are some difficulties when rounding floating point numbers.
See the help page of the method `awayfromzero()`.

Rounding can be chained by pipe operator `|>`.

```{r}
totalize(datasets::CO2, row=c(Treatment, conc), col=Type, val=uptake, FUN=mean) |>
    awayfromzero(1)
```


