---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# totalize

<!-- badges: start -->
<!-- badges: end -->

[Japanese README is here.](README-ja.md)

Aggregate values with its group total and put into matrix tables

## Installation

You can install `totalize` from GitHub.

``` r
devtools::install_github("ichiromurata/totalize")
```

## How this works

Summarising data can be acted by `table()`,

```{r}
table(interaction(datasets::CO2$Type, datasets::CO2$Treatment))
```

or `by()`,

```{r}
by(datasets::CO2$uptake, datasets::CO2[c("Type", "Treatment")], sum)
```

or `stats::aggregate()`.

```{r}
stats::aggregate(datasets::CO2["uptake"], datasets::CO2[c("Type", "Treatment")], sum)
```

But if you need subtotals and total (total "Type", total "Treatment" and both summed in examples above), 
you have to calculate them through several steps, since it involves multiple count of the same data essentially 
and can't be done by one method execution.

`totalize` rolls up these operations.

```{r}
library(totalize)
totalize(datasets::CO2, "uptake", c("Type", "Treatment"), sum) |> toDF()
```

Further, it provides matrix representation.

```{r}
totalize(datasets::CO2, "uptake", c("Type", "Treatment"), sum) |> tomatrix(1)
```

3 or more variables can be put into a (2d) matrix.

```{r}
totalize(datasets::CO2, "uptake", c("Type", "Treatment", "conc"), sum) |> tomatrix(1)

totalize(datasets::CO2, "uptake", c("Type", "Treatment", "conc"), sum) |> tomatrix(c(1, 3))
```

Would you like to ignore NAs? Pass `na.rm=TRUE` to the arguments.

```{r}
# Missing values turn the totals into NA
within(datasets::CO2, uptake[c(1, 10)] <- NA) |>
    totalize("uptake", c("Type", "Treatment", "conc"), sum) |> tomatrix(1)

# Ignoring NAs
within(datasets::CO2, uptake[c(1, 10)] <- NA) |>
    totalize("uptake", c("Type", "Treatment", "conc"), sum, na.rm=TRUE) |> tomatrix(1)
```

## Rounding numbers

This package includes a rounding method which rounds half [away from zero](https://en.wikipedia.org/wiki/Rounding#Round_half_away_from_zero).

There are some difficulties when rounding floating point numbers.
See the help page of the method `awayfromzero()`.

Rounding can be chained by pipe operator `|>`.

```{r}
totalize(datasets::CO2, "uptake", c("Type", "Treatment", "conc"), mean) |>
    tomatrix(c(2, 3)) |> awayfromzero(1)
```


